Summary of cell type ontologies in reference files

### 2024-12-20 analyses/cell-type-consensus/exploratory-notebooks/01-reference-exploration.Rmd
fix module typos and re-render notebook. notebook was also styled by precommit

```{r}
# make density plots showing distribution of ancestors and descendants 
ggplot(cl_df, aes(x = total_ancestors)) +
  geom_density(fill = "#00274C", alpha = 0.5) +
  geom_vline(data = ancestor_labels_df,
             mapping = aes(xintercept = total_ancestors), 
             lty = 2) +
  geom_text(
    data = ancestor_labels_df,
    mapping = aes(x = total_ancestors, y = 0.04, label = cl_annotation),
    angle = 90, 
    vjust = -0.5
  ) +
  labs(
    x = "Number of ancestors",
    y = "Density"
  )
```

@@

```{r}
ggplot(cl_df, aes(x = total_descendants)) +
  geom_density(fill = "#FFCB05", alpha = 0.5) +
  geom_vline(data = line_df,
             mapping = aes(xintercept = total_descendants), 
             lty = 2) +
  geom_text(
    data = line_df,
    mapping = aes(x = total_descendants, y = 0.6, label = cl_annotation),
    angle = 90, 
    vjust = -0.5
  ) +
  labs(
    x = "Number of descendants",
    y = "Density"
  )
```

@@

```{r}
ggplot(cl_df, aes(x = total_descendants)) +
  geom_density(fill = "#FFCB05", alpha = 0.5) +
  geom_vline(data = line_df,
             mapping = aes(xintercept = total_descendants), 
             lty = 2) +
  geom_text(
    data = line_df,
    mapping = aes(x = total_descendants, y = 0.6, label = cl_annotation),
    angle = 90, 
    vjust = -0.5
  ) +
  labs(
    x = "Number of descendants",
    y = "Density"
  ) +
  xlim(c(0,500))
```

@@

```{r}
# get a data frame with all combinations of panglao and blueprint terms
# one row for each combination 
all_ref_df <- expand.grid(panglao_df$panglao_ontology,
                          blueprint_df$blueprint_ontology) |> 
  dplyr::rename(
    panglao_ontology = "Var1",
    blueprint_ontology = "Var2"
  ) |> 
  # add in the human readable values for each ontology term
  dplyr::left_join(blueprint_df, by = "blueprint_ontology") |> 
  dplyr::left_join(panglao_df, by = "panglao_ontology") |> 
  tidyr::drop_na() |> 
  dplyr::rowwise() |> 
  dplyr::mutate(
    # least common shared ancestor
    lca = list(rownames(ontoProc::findCommonAncestors(blueprint_ontology, panglao_ontology, g = cl_graph)))
  )

@@

```{r}
ggplot(lca_df, aes(x = total_ancestors)) +
  geom_density() +
  geom_vline(data = ancestor_labels_df,
             mapping = aes(xintercept = total_ancestors), 
             lty = 2) +
  geom_text(
    data = ancestor_labels_df,
    mapping = aes(x = total_ancestors, y = 0.6, label = cl_annotation),
    angle = 90, 
    vjust = -0.5
  ) +
  labs(
    x = "Total number of ancestors",
    y = "Density"
  )
```

```{r}
ggplot(lca_df, aes(x = total_descendants)) +
  geom_density() +
  geom_vline(data = line_df,
             mapping = aes(xintercept = total_descendants), 
             lty = 2) +
  geom_text(
    data = line_df,
    mapping = aes(x = total_descendants, y = 0.002, label = cl_annotation),
    angle = 90, 
    vjust = -0.5
  ) +
  labs(
    x = "Total number of descendants",
    y = "Density"
  )
```

Let's zoom into the area below 1000, since we already know we would want to exlude anything above that based on this plot.

```{r}
ggplot(lca_df, aes(x = total_descendants)) +
  geom_density() +
  geom_vline(data = line_df,
             mapping = aes(xintercept = total_descendants), 
             lty = 2) +
  geom_text(
    data = line_df,
    mapping = aes(x = total_descendants, y = 0.002, label = cl_annotation),
    angle = 90, 
    vjust = -0.5
  ) +
  xlim(c(0, 1000)) +
  labs(
    x = "Total number of descendants",
    y = "Density"
  )
```

@@

We can also look at all the cell types we are keeping and the total number of descendants to see if there are any that may be we don't want to include because the term is too braod.

@@

```{r}
# progenitor cell
print_df |> 
  dplyr::filter(cl_annotation == "progenitor cell") |> 
  head(n=15) # there's a lot of these so let's only print out some 
```

@@

# get similarity index for each set of terms 
si_df <- lca_df |> 
  dplyr::rowwise() |> 
  dplyr::mutate(
    similarity_index = ontologySimilarity::get_sim_grid(ontology = cl_ont,
                                                        term_sets = list(panglao_ontology, blueprint_ontology)) |> 
      ontologySimilarity::get_sim()
  )
```

@@

    ggplot(si_df, aes(x = similarity_index)) +
      geom_density() +
      geom_vline(data = line_df,
                 mapping = aes(xintercept = similarity_index), 
                 lty = 2) +
      labs(
        x = "Similarity index",
        y = "Density",
        title = celltype
      )

  })
```

>>>>

```{r}
# make density plots showing distribution of ancestors and descendants
ggplot(cl_df, aes(x = total_ancestors)) +
  geom_density(fill = "#00274C", alpha = 0.5) +
  geom_vline(
    data = ancestor_labels_df,
    mapping = aes(xintercept = total_ancestors),
    lty = 2
  ) +
  geom_text(
    data = ancestor_labels_df,
    mapping = aes(x = total_ancestors, y = 0.04, label = cl_annotation),
    angle = 90,
    vjust = -0.5
  ) +
  labs(
    x = "Number of ancestors",
    y = "Density"
  )
```

@@

```{r}
ggplot(cl_df, aes(x = total_descendants)) +
  geom_density(fill = "#FFCB05", alpha = 0.5) +
  geom_vline(
    data = line_df,
    mapping = aes(xintercept = total_descendants),
    lty = 2
  ) +
  geom_text(
    data = line_df,
    mapping = aes(x = total_descendants, y = 0.6, label = cl_annotation),
    angle = 90,
    vjust = -0.5
  ) +
  labs(
    x = "Number of descendants",
    y = "Density"
  )
```

@@

```{r}
ggplot(cl_df, aes(x = total_descendants)) +
  geom_density(fill = "#FFCB05", alpha = 0.5) +
  geom_vline(
    data = line_df,
    mapping = aes(xintercept = total_descendants),
    lty = 2
  ) +
  geom_text(
    data = line_df,
    mapping = aes(x = total_descendants, y = 0.6, label = cl_annotation),
    angle = 90,
    vjust = -0.5
  ) +
  labs(
    x = "Number of descendants",
    y = "Density"
  ) +
  xlim(c(0, 500))
```

@@

```{r}
# get a data frame with all combinations of panglao and blueprint terms
# one row for each combination
all_ref_df <- expand.grid(
  panglao_df$panglao_ontology,
  blueprint_df$blueprint_ontology
) |>
  dplyr::rename(
    panglao_ontology = "Var1",
    blueprint_ontology = "Var2"
  ) |>
  # add in the human readable values for each ontology term
  dplyr::left_join(blueprint_df, by = "blueprint_ontology") |>
  dplyr::left_join(panglao_df, by = "panglao_ontology") |>
  tidyr::drop_na() |>
  dplyr::rowwise() |>
  dplyr::mutate(
    # least common shared ancestor
    lca = list(rownames(ontoProc::findCommonAncestors(blueprint_ontology, panglao_ontology, g = cl_graph)))
  )

@@

```{r}
ggplot(lca_df, aes(x = total_ancestors)) +
  geom_density() +
  geom_vline(
    data = ancestor_labels_df,
    mapping = aes(xintercept = total_ancestors),
    lty = 2
  ) +
  geom_text(
    data = ancestor_labels_df,
    mapping = aes(x = total_ancestors, y = 0.6, label = cl_annotation),
    angle = 90,
    vjust = -0.5
  ) +
  labs(
    x = "Total number of ancestors",
    y = "Density"
  )
```

```{r}
ggplot(lca_df, aes(x = total_descendants)) +
  geom_density() +
  geom_vline(
    data = line_df,
    mapping = aes(xintercept = total_descendants),
    lty = 2
  ) +
  geom_text(
    data = line_df,
    mapping = aes(x = total_descendants, y = 0.002, label = cl_annotation),
    angle = 90,
    vjust = -0.5
  ) +
  labs(
    x = "Total number of descendants",
    y = "Density"
  )
```

Let's zoom into the area below 1000, since we already know we would want to exclude anything above that based on this plot.

```{r}
ggplot(lca_df, aes(x = total_descendants)) +
  geom_density() +
  geom_vline(
    data = line_df,
    mapping = aes(xintercept = total_descendants),
    lty = 2
  ) +
  geom_text(
    data = line_df,
    mapping = aes(x = total_descendants, y = 0.002, label = cl_annotation),
    angle = 90,
    vjust = -0.5
  ) +
  xlim(c(0, 1000)) +
  labs(
    x = "Total number of descendants",
    y = "Density"
  )
```

@@

We can also look at all the cell types we are keeping and the total number of descendants to see if there are any that may be we don't want to include because the term is too broad.

@@

```{r}
# progenitor cell
print_df |>
  dplyr::filter(cl_annotation == "progenitor cell") |>
  head(n = 15) # there's a lot of these so let's only print out some
```

@@

# get similarity index for each set of terms
si_df <- lca_df |>
  dplyr::rowwise() |>
  dplyr::mutate(
    similarity_index = ontologySimilarity::get_sim_grid(
      ontology = cl_ont,
      term_sets = list(panglao_ontology, blueprint_ontology)
    ) |>
      ontologySimilarity::get_sim()
  )
```

@@

    ggplot(si_df, aes(x = similarity_index)) +
      geom_density() +
      geom_vline(
        data = line_df,
        mapping = aes(xintercept = similarity_index),
        lty = 2
      ) +
      labs(
        x = "Similarity index",
        y = "Density",
        title = celltype
      )
  })
```
